---
title: "cestode toy data1"
author: "brian avery"
date: "July 7, 2016"
output: html_document
---
```{r echo=FALSE}
options(width = 99)
## this is a bunch of formatting stuff that I like echo=FALSE means that it doesn't show in the Rmd
## uncomment the next line if the htmlTable package hasn't been installed yet, only need to do it once
# install.packages("htmlTable")
library(htmlTable)
```
## Intro
Making some test/toy data for Syd to use to start learning/doing some data analysis in R while she gathers data. It's not the best toy data ever, but it's a start.  

### Important considerations:
<ol><li>we have different sample collection sites
<li>data was collected over several years
<li>each site should have a mix of infected and uninfected results
</ol>

### Cestode data code book
(this test data is also a nice test of the format of the dataset so for now this should be considered a draft)  
in general in a code book, each column corresponds to a variable, the code book has the following info for each column.  
this is the format we'll use:
<li>Column label: the name of the varialbe in the dataset
<li>Variable: an explanation of what the variable is/measures
<li>Variable type: numeric, integer, character, logical (Boolean), factor, date, etc.
<li>Allowable values: list of possible values
<li>Comments: anything else

**year**
<li>Column label: year
<li>Variable: year individual was collected
<li>Variable type: integer
<li>Allowable values: 2010-2016
<li>Comments:  

**site**
<li>Column label: site
<li>Variable: site from whence individual was collected
<li>Variable type: character
<li>Allowable values: blackrock, statemarina, antelopeisland, sprialjetty
<li>Comments: 

**ID**
<li>Column label: ID
<li>Variable: indentifier of individual, a few letters that identify the site
<li>Variable type: character
<li>Allowable values: BR, SM, AI, or SJ followed by a number from 1-200
<li>Comments: BR=blackrock, SM=statemarina, AI=antelopeisland, SJ=sprialjetty. numbers are individual number within that site

**infected**
<li>Column label: infected
<li>Variable: infection status of individual
<li>Variable type: logical
<li>Allowable values: TRUE, FALSE
<li>Comments: TRUE=infected, FALSE=uninfected. this is of course our best call based on the PCR

## Generate the toy data

First we have a lot of numbers etc. to make, then we want to format it into a data frame to make it easy to work with.

```{r}
# this ensures that the random numbers generated are the same if we re-run the code later
set.seed(84105)

## build each column, the ID col is complicated so it gets built in 2 pieces here and combined later
years <- rep(c(2012, 2013), each = 120)
sites <- rep(c("blackrock", "statemarina", "antelopeisland", "sprialjetty"), each = 30, times = 2)
IDL <- rep(c("BR", "SM", "AI", "SJ"), each = 30, times = 2)
IDno <- rep(c(1:30), 8)

## get some infection data, generated from four differently weighted distributions
## so the different sites might have different outcomes
inf1a <- sample(c(TRUE, FALSE, NA), 30, replace = TRUE, prob = c(.49, .49, 0.02))
inf1b <- sample(c(TRUE, FALSE, NA), 30, replace = TRUE, prob = c(.49, .49, 0.02))
inf1c <- sample(c(TRUE, FALSE, NA), 30, replace = TRUE, prob = c(.49, .49, 0.02))
inf1d <- sample(c(TRUE, FALSE, NA), 30, replace = TRUE, prob = c(.49, .49, 0.02))
inf2 <- sample(c(TRUE, FALSE, NA), 30, replace = TRUE, prob = c(.74, .24, 0.02))
inf3a <- sample(c(TRUE, FALSE, NA), 30, replace = TRUE, prob = c(.44, .54, 0.02))
inf3b <- sample(c(TRUE, FALSE, NA), 30, replace = TRUE, prob = c(.44, .54, 0.02))
inf4 <- sample(c(TRUE, FALSE, NA), 30, replace = TRUE, prob = c(.94, .04, 0.02))

## combine into one vector for the infected column
infected <- c(inf1a, inf1b, inf3a, inf2, inf1c, inf1d, inf3b, inf4)
```

The `dplyr` and `tidyr` packages are really good for working with data frames. Look them up. Here I'm going to use them to format the numbers generated above into a nice, tidy data frame.

```{r message=FALSE}
# loads the packages so we can use them, 
# you need to use install.packages() if you don't have them yet
library(dplyr)
library(tidyr)
```

```{r}
## build the start of the dataframe out of the first few simulated columns of data
cestodetoy <- data.frame(year=years, site=sites, IDL=IDL, IDno=IDno)

## combine the 2 pieces of the ID column into 1 piece and remove the old halves
cestodetoy <- unite(cestodetoy, ID, IDL, IDno, sep = '', remove = TRUE)

## add the infected column
cestodetoy$infected <- infected

```

Now that the entire data set is built, let's look at it.

```{r}
# 3 really good, quick ways to see what the data looks like.
head(cestodetoy)
str(cestodetoy)
summary(cestodetoy)
```

`dplyr` is your new and best-est friend for summarzing data!  
the pipe symbol `%>%` is super helpful in using `dplyr` for summarizing data since you can string together multiple commands.
another interesting/weird thing is that `TRUE=1` and `FALSE=0` so you can sum and mean them!

```{r}
# grouped by site (not years) the "na.rm=TRUE" removes NAs if there are any
cestodetoy %>% group_by(site) %>% summarise(sum(infected, na.rm=TRUE))

## the mean of infected gives you proportion
cestodetoy %>% group_by(site) %>% summarise(mean(infected, na.rm=TRUE))

## grouped by site and then year
cestodetoy %>% group_by(site, year) %>% summarise(sum(infected, na.rm=TRUE))
cestodetoy %>% group_by(site, year) %>% summarise(mean(infected, na.rm=TRUE))
```

```{r}
## write to csv file
## uncomment the next line if you want to write the data to a csv file

# write.csv(cestodetoy, file = "cestode_toy_dataset1.csv", row.names=FALSE)
```

Now you can use ggplot2 to make some plots and look at the data. I'll start you off with one simple graph.

```{r}
# loads the package so we can use it, 
# you need to use install.packages() if you don't have it yet
library(ggplot2)

# a simple histogram of the number of infected animals per site
# should really be proportion, but in this fake example all of the sample sizes are the same

# recreate the summarized data from above, but this time assign it to allinfected
allinfected <- cestodetoy %>% group_by(site) %>% summarise(sum(infected, na.rm=TRUE))
allinfected

# the title of the second column that contains the sum of the infected has a terrible name
# let's fix that
names(allinfected)[names(allinfected) == "sum(infected, na.rm = TRUE)"] <- "number_infected"
allinfected

# make the histogram
ggplot(data=allinfected, aes(x= site, y=number_infected)) +
  geom_bar(stat="identity") +
  theme_bw() 
```

your turn!
